### 2025-05-21 docs/article/Sections/state-of-the-art.tex
chapter 2 - Explain Supported Data Models APIs

On the other hand, \textit{internal} DSLs for HTML allow templates to be defined 
directly within the \emph{host} language (such as Java, Kotlin, Groovy, Scala, or 
other general-purpose programming languages), rather than using text-based 
template files~\cite{carvalho2019hot}. In this case, a web template is not limited 
to templating constructs but may instead leverage any available primitive of the 
host language or any external API.
For example, in Java, you may iterate over data using a \texttt{for} statement,
or using the \texttt{java.util.streams} API, or any other JVM library such as
Guava, Varv, StreamEx, or others—according to your preferences.

@@

Restricted or unrestricted.

>>>>

On the other hand, \textit{internal} DSLs for HTML allow templates to be defined 
directly within the \emph{host} language (such as Java, Kotlin, Groovy, Scala, or 
other general-purpose programming languages), rather than using text-based 
template files~\cite{carvalho2019hot}. In this case, a web template is not limited 
to templating constructs but may instead leverage any available primitive of the 
host language or any external API.

@@

Template engines with dedicated templating dialects typically rely on specific 
control flow constructs—such as a \texttt{foreach}-like statement—to iterate 
over elements in a data source. In the Java ecosystem, \texttt{Iterable} is the 
common supertype implemented by most data structures and serves as the standard 
API for iteration. However, other standard interfaces, such as 
\texttt{java.util.Stream}, also represent sequences of elements but are not 
compatible with \texttt{Iterable}.
This fragmentation becomes more evident when considering other JVM languages, 
such as Kotlin, which introduces additional abstractions like the 
\texttt{Sequence} interface—also incompatible with \texttt{Iterable}. As a result, 
template engines based on \textit{external} DSLs must explicitly support each of these 
interfaces to accommodate the various iteration protocols across languages and 
libraries.

On the other hand, template engines that employ an \textit{internal} DSL do not 
face such limitations, as they can leverage any available API within the host 
programming language. This flexibility extends beyond the standard library to 
include third-party APIs as well.
For example, in Java, developers can iterate over data using a traditional 
\texttt{for} loop, the \texttt{java.util.stream} API, or external libraries such 
as Guava, Vavr, StreamEx, and others—depending on their specific needs or preferences.
